Prominent claims of group polarization are plausibly false with lessons for studying psychological dynamics

### 2024-04-09 GroupPolarizationStatmod/paper.bk/ms.tex
init quarto paper rewrite

% % Author: Matthew Turner

% % \documentclass[11pt,letterpaper]{scrartcl}
% % \documentclass[letterpaper,man,natbib]{apa6}
% \documentclass[11pt,letterpaper]{amsart}
% % \documentclass[11pt, letterpaper]{article}
% % \usepackage{natbib}
% % \documentclass[11pt]{report}
% % \documentclass{report}
% % \documentclass{book}
% \usepackage[bookmarks, hidelinks]{hyperref}
% \usepackage{amssymb,amsmath}
% % \usepackage{fullpage}
% \usepackage{tabulary}
% \usepackage{tabularx}
% \usepackage{float}
% % \usepackage[margin=1.00in]{geometry}
% \usepackage[margin=0.90in]{geometry}

% % \usepackage{fontspec}

% % \usepackage{xunicode}
% % \usepackage{xltxtra}
% % \defaultfontfeatures{Mapping=tex-text}
% % \setmainfont[Ligatures={Common}, Numbers={OldStyle}, Variant=01]{Hoefler Text}
% % \setromanfont[Ligatures={Common}, Numbers={OldStyle}, Variant=01]{Hoefler Text}
% % \setmonofont[Scale=0.95]{PT Sans}
% \usepackage{caption}
% \usepackage{apacite}
% \usepackage{booktabs}
% \usepackage{pslatex}
% \usepackage{subcaption}
% \usepackage{pgfplots}
% \usepackage{wrapfig}
% \usepackage[english]{babel}
% \usepackage{lmodern}
% % \doublespace
% % \usepackage{url}
% \usepackage{bigfoot}
% \usepackage[export]{adjustbox}
% % \setlength\intextsep{0pt}

% \usepackage{graphicx}

\documentclass[12pt, letterpaper]{article}
\usepackage{fontspec} 
\usepackage{marvosym}
\usepackage{setspace}
\usepackage{longtable}
% DOCUMENT LAYOUT
% \usepackage{geometry} 
\usepackage[margin=0.90in]{geometry}
\usepackage{lineno}
\usepackage[usenames,dvipsnames]{xcolor}
% \geometry{letterpaper, textwidth=5.5in, textheight=8.5in, marginparsep=7pt, marginparwidth=.6in}
% \setlength\parindent{0in}

@@

% \renewcommand{\thesubsection}{\arabic{subsection}.}
% FONTS
% \usepackage[usenames,dvipsnames]{color}
% \usepackage{xunicode}
% \usepackage{xltxtra}
\defaultfontfeatures{Mapping=tex-text}
% \setromanfont[Ligatures={Common}, Variant=01]{Hoefler Text}
% \setromanfont[Ligatures={Common}, Variant=01]{Palatino}
\setromanfont[Ligatures={Common}, Variant=01]{Baskerville}
% \setromanfont[Ligatures={Common}, Variant=01]{Plantagenet Cherokee}
% \setmainfont [BoldFont={STIX Two Text SemiBold}, Ligatures={Common},
% Numbers={OldStyle}, Variant=01]{STIX Two Text}
% \setmainfont [BoldFont={STIX Two Text SemiBold}, Ligatures={Common},
% Numbers={OldStyle}, Variant=01]{STIX Two Text}
% \setromanfont [Ligatures={Common},
% Numbers={OldStyle}, Variant=01]{STIX Two Text}
\setmonofont[Scale=0.95]{PT Sans}

% ---- CUSTOM COMMANDS
% \chardef\&="E050
\newcommand{\html}[1]{\href{#1}{\scriptsize\textsc{[html]}}}
\newcommand{\pdf}[1]{\href{#1}{\scriptsize\textsc{[pdf]}}}
\newcommand{\doi}[1]{\href{#1}{\scriptsize\textsc{[doi]}}}
\newcommand{\lurl}[1]{\href{#1}{\scriptsize\textsc{[link]}}}

\newcommand{\githubsymbol}{\raisebox{-2pt}{\includegraphics[height=10.5pt]{github}}}
\newcommand{\twittersymbol}{\raisebox{-2pt}{\includegraphics[height=10.5pt]{twitter}}}
\newcommand{\monitorsymbol}{\raisebox{-2pt}{\includegraphics[height=10.5pt]{monitor}}}

% ---- MARGIN YEARS
\usepackage{marginnote}
\newcommand{\amper{}}{\chardef\amper="E0BD }
\newcommand{\years}[1]{\marginnote{\small #1}}
\renewcommand*{\raggedleftmarginnote}{}
\setlength{\marginparsep}{7pt}
\reversemarginpar

% HEADINGS
\usepackage{sectsty} 
\usepackage[normalem]{ulem} 
\sectionfont{\upshape\Large}
\subsectionfont{\scshape\normalsize} 
\subsubsectionfont{\upshape\large}

% PDF SETUP
% ---- FILL IN HERE THE DOC TITLE AND AUTHOR
\usepackage[%dvipdfm, 
bookmarks, colorlinks, breaklinks, 
% ---- FILL IN HERE THE TITLE AND AUTHOR
	pdftitle={Matthew A. Turner - Vita},
	pdfauthor={Matthew A. Turner, PhD},
	pdfproducer={https://mt.digital}
]{hyperref}  
\hypersetup{linkcolor=blue,citecolor=blue,filecolor=black,urlcolor=MidnightBlue}

@@

@@

\title{Group polarization is often a mirage caused by inappropriate statistical analyses}
% \shorttitle{Group polarization or simple consensus?}
\author[1,2]{{Matthew A.~Turner}}
\affil[1]{Department of Earth System Science, Stanford University}
\affil[2]{Division of Social Sciences, Stanford Doerr School of Sustainability, Stanford University}

\author[3,4,5]{{Paul E.~Smaldino}}
\affil[3]{Cognitive and Information Sciences, University of California, Merced}
\affil[4]{Center for Advanced Study in the Behavioral Sciences, Stanford University}
\affil[5]{Santa Fe Institute}

@@

\linenumbers
\modulolinenumbers[5]

\begin{abstract}
  \noindent
  Group polarization is the observed tendency of
  like-minded groups to arrive at a more extreme consensus opinion than
  the average opinion after discussing some topic of interest.
  Clearly it is important to understand such a phenomenon to understand broader
  problematic trends towards increased polariztion across society as a whole.
  Unfortunately, recent analyses have demonstrated that this measurement and
  analysis approach can result in false positive detections group polarization
  from failing to account for floor/ceiling effects induced when using ordinal scales to
  measure continuously-distributed latent psychological variables. 
  Here we demonstrate that many high profile detections
  of group polarization are plausibly false
  due to these floor/ceiling effects using generative modeling seeded by 
  reported group polarization values and parameters. After we identify which
  reported detections of group polarization are plausibly false detections, we further
  investigate whether the latent constant mean opinion and pre- and post-discussion
  opinion variances can fool the statistical tests used in each study to 
  generate false positives. Finally, we evaluate whether a more appropriate
  Bayesian cumulative normal statistical model can more accurately distinguish
  group polarization from simple consensus. Group polarization continues to 
  pose a problem for psychological theory with an ever-expanding set of potential theoretical
  explanations. A search for explanations is pointless, however, without sound
  evidence. We believe this work will provide a sturdier analytical foundation for 
  future group-polarization research.
\end{abstract}

@@

@@

@@

\section{Model and Data}

Our model represnts the generative process of finding different forms of group consensus
in group polarization experiments, and the process of measuring that consensus using
an ordinal measurement scale, e.g.\ a Likert-type scale. This model considers two
relevant forms of consensus: \emph{simple consensus} and \emph{group polarization}.
Either process is \emph{consensus} since opinion variance among the group 
decreases as individuals come to agree more than they did previously. 
Group polarization experiments are designed to foster group consensus
it is not necessary to consider clustering, negative influence, or other
processes known to generally occur when opinions within a group significantly
diverge~\cite{Turner2018}.

We collected group polarization data from ten empirical studies with sixty constituent
experimental treatments across all studies, specifically the opinion shift and pre- and
post-deliberation mean opinions, and the number of participants
in each treatment (\citeA{Burnstein1975} did not report pre- and post-deliberation
means, only the shift in means). We did not include experimental treatments that did not 
yield a positive detection of group polarization since we are only estimating
the rate of plausible false positive detections of group polarization. 
We then created new data that indicated whether a reported.

In two cases, reported opinion and shift values were converted from an ordinal
scale to a percentage scale, specifically \citeA{Friedkin1999a} and \citeA{Krizan2007}.
The authors in these papers did report the original ordinal measurement scale,
so we converted the means back to the orignial measurement scale.
In a few cases from \citeA{Friedkin1999a} we did not need to apply our 
model to the reported data because we could determine a reported opinion shift was 
\emph{prima facie} false since reported opinion shift magnitudes
were smaller than the standard deviation in shifts (see Table 1, p. 868, \emph{ibid}),
despite Friedkin's claim that regressions (not clearly defined) on all shifts 
yielded $p < 0.05$~\cite{Kruschke2018c}.

\subsection{Latent opinions, ordinal measurements}

Our analysis relies on two important formal concepts: latent opinions and their 
measurement on an ordinal scale. \emph{Latent opinions} are unobservable psychological,
continuous-valued variables that are somehow represented and stored in the
brain-body. 
When researchers use continuous statistical models to infer 
that group polarization occurred (which all ten considered here do), 
they are implicitly assuming that latent opinions are themselves continuous.
We can only measure behaviors in group polarization experiments,
specifically the behavior of participants marking one ordinal value or another
to indicate their opinion---a process that is not well understood in itself,
but which is unimportant for our analyses. \emph{Ordinal measurements} (e.g.\
a Likert or other discrete-valued opinion scale) of latent opinions force
participants to appropriately bin their opinion in one of several categories
representing the valence and extremity of their opinion, e.g., on a
7-point Likert scale -3 could indcate
``strongly disagree'', +3 could be ``strongly agree'', 0 would be neutral, and 
intermediate values are less extreme opinions of either valence.

All ten studies analyzed here used an ordinal scale, but group polarization experiments seem to
lack any systematicity in experimental design beyond that. (EXAMPLES)

\subsection{Simple consensus and group polarization are statistical patterns}

\subsection{Analytical approach}

The goal of this study is to understand if published detections of group polarization are
plausibly false.

Therefore we designed a series of tests to understand if, and to what degree,
published results are false positive detections of group polarization that
are at least as well described by a simple-consensus model.

\subsubsection{Identifying parameters that plausibly generate false detections of group polarization}

\subsubsection{$t$-tests of generated simple-consensus data}

\subsubsection{Bayesian ordered-probit models of generated simple-consensus data}

@@

\spacing{1.0}
\input{plausibleFP-table.tex}
\spacing{1.5}

\subsection{Do $t$-tests accurately detect simple consensus?}

We found that in X\% of generated simple consensus data led to false positive
detections

When simulated data reliably result in false positive $t$-test results (where
the $t$-test $p \le 0.1$) this strengthens the case that reported detections
of a false group polarization detection.

\spacing{1.0}
% \input{t-test-table.tex}
\input{t-test-table-custom.tex}
\spacing{1.5}

\subsection{Do Bayesian ordered-probit models accurately detect simple consensus?}

@@

@@

This problem is made all the worse by the lack of a consistent experimental design for group
polarization experiments. This makes it impossible to systematically account for
additional sources of vairance known to lead to (orthogonal) over-estimates
of effect sizes~\cite{Yarkoni2021}. This is due to the rush to explain why 
group polarization occurs using weak theory~\cite{Cartwright1971,Cartwright1973} and makes
group polarization meta-analyses (e.g. Brown, 1986; and Isenberg, 1986)\nocite{Brown1986,Isenberg1986}
unreliable~\cite{Meehl1990,Meehl1997}.

>>>>

\documentclass[11pt, letterpaper]{article}
\usepackage{marvosym}
\usepackage{float}
\usepackage{setspace}
\usepackage{longtable}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{pdflscape}
\usepackage{graphicx}
% DOCUMENT LAYOUT
% \usepackage{geometry} 
\usepackage[margin=1.0in]{geometry}
\usepackage{lineno}
\usepackage[usenames,dvipsnames]{xcolor}
\definecolor{myorange}{RGB}{240, 96, 0}
\newcommand{\mt}[1]{{\textcolor{myorange} {({\tiny MT:} #1)}}}
% \geometry{letterpaper, textwidth=5.5in, textheight=8.5in, marginparsep=7pt, marginparwidth=.6in}
% \setlength\parindent{0in}

@@

\usepackage{amsmath}
\usepackage{mathspec}
% \setallmainfonts(Digits,Latin){STIX Two Text}
% \setallmainfonts(Digits,Latin){Baskerville}
% \setallmainfonts(Digits,Latin){Plantagenet Cherokee}
\setallmainfonts(Digits,Latin){Palatino}
% \setallmainfonts(Digits,Latin){Hoefler Text}

% HEADINGS
% \usepackage{sectsty} 
% \usepackage[normalem]{ulem} 
% \sectionfont{\upshape\Large}
% \subsectionfont{\scshape\normalsize} 
% \subsubsectionfont{\upshape\large}

% PDF SETUP
% ---- FILL IN HERE THE DOC TITLE AND AUTHOR
\usepackage[%dvipdfm, 
bookmarks, colorlinks, breaklinks, 
% ---- FILL IN HERE THE TITLE AND AUTHOR
	pdftitle={Many claims of group polarization are plausibly false due to inappropriate
    statistics},
	pdfauthor={Matthew A. Turner},
	pdfproducer={https://mt.digital}
]{hyperref}  
\hypersetup{linkcolor=blue,citecolor=blue,filecolor=black,urlcolor=MidnightBlue}

@@

\newcommand{\sigmapre}{\sigma_\mathrm{pre}}
\newcommand{\sigmapost}{\sigma_\mathrm{post}}

\newcommand{\normalpre}{\mathcal{N}(\mu, \sigmapre)}
\newcommand{\normalpost}{\mathcal{N}(\mu, \sigmapost)}

\newcommand{\ordmeanpre}{\langle o_{\mathrm{pre}}}
\newcommand{\ordmeanpost}{\langle o_{\mathrm{post}}}

@@

\title{Many claims of group polarization are plausibly false due to inappropriate
    statistics}
\author[1]{{Matthew A.~Turner}}
\affil[1]{Division of Social Sciences, Stanford Doerr School of Sustainability, Stanford University}

\author[2,3]{{Paul E.~Smaldino}}
\affil[2]{Cognitive and Information Sciences, University of California, Merced}
\affil[3]{Santa Fe Institute}

\date{}

@@

\linenumbers
% \modulolinenumbers[3]

\begin{abstract}
  \noindent
  \emph{Group polarization} is a theoretical process where like-minded novel group members 
  become more extreme in their opinions after group discussion, but also find \emph{consensus}, reducing 
  group-level opinion variance.
  \emph{Simple consensus} may occur instead, where opinion variance decreases but
  the mean group opinion does not change.
  Here we show that many published detections of group polarization are explained
  equally rigorously as simple consensus.
  This is the case because experimenters failed
  to account for the measurement of continous-valued psychological opinions 
  as ordinal-valued reported opinions, e.g., on a Likert-type scale. 
  This creates a ceiling effect that masks extreme opinions becoming less extreme.
  We developed a generative model to 
  demonstrate that 95\% of sixty experimental conditions
  across ten group polarization papers report plausibly false detections of
  group polarization. Our approach can be used to evaluate whether other detections of
  group polarization may be plausibly false---our approach could be applied beyond
  group polarization as well. This work also provides a
  framework for new group polarization studies that could
  share a common, rigorous, and interoperable study design to multiply our knowledge
  of this potentially important group process, avoiding the historically fragmented
  approach using disparate, incommensurate study designs.
\end{abstract}

\clearpage

\begin{quote}
In our introductory social psychology course, 
we have for many years used the [group polarization experimental paradigm] as
a laboratory exercise. The exercise works beautifully, but one must be
careful to forewarn a class that [group polarization] does not occur with every 
group\ldots and that the effect is not large. 
\par\raggedleft\cite[p. 205]{Brown1986}
\end{quote}

\begin{quote}
One of the most robust findings in social psychology is that of attitude polarization 
following discussion with like-minded others.
\par\raggedleft\cite[p. 267]{Cooper2001}
\end{quote}

@@

\begin{itemize}
  \item 
    \emph{Group polarization} is a process where the mean opinion of a 
    like-minded group becomes more extreme over time~\cite{Brown1986,Brown2000}. 
    It is therefore an important component process in
    the broader process of increasing political polarization that can hamstring
    governments~\cite{Mason2018UncivilAgreementBook,Klein2020}.
  \item
    Individuals' opinions in group polarization experiments are typically 
    measured using an ordinal scale, such
    as a Likert scale, where behavioral experiment participants' internal, theoretically
    continuous, opinions are converted to one of several opinion bins indiciating
    their degree of agreement or disagreement with some prompt statement.
  \item
    Recently, however, it has been shown that it is critical to use 
    appropriate statistical methods when using experimental designs that 
    measure continuous latent psychological variables with ordinal
    measurements---failure to do so can result in false positive support for
    the presence of experimental effects~\cite{Liddell2018}.
  \item
    In this paper, we develop a generative modeling approach for evaluating whether or
    not published detections of group polarization are plausibly false, i.e.,
    whether they are \emph{plausibly-false positives} (PFPs). 
  \item
    We found significant evidence that many published group polarization findings
    are indeed plausibly false since we found distribution parameters that
    generated simulated pre- and post-discussion data that a $t$-test found
    to be from distributions with different means, although there was no difference
    in mean, only a difference in variance. 
  \item
    Although a $t$-test may be fooled by simple consensus under ordinal measurements,
    we show that a more appropriate ordered probit model of the data yield no
    such plausibly-false positives.
  \item
    Our work shows the need for using appropriate statistical tests, as well as
    the need for a consistent experimental paradigm so that meaningful meta-analyses
    can be performed for theoretical integration and rigorous understanding of
    the group polarization effect, should it prove to occur in future group
    polarization studies that use proper statistics. In addition to avoiding
    $t$-tests, future group polarization studies can be more reliable by
    accounting for random effects arising from, e.g., socioeconomic status,
    gender, and other individual-level factors that may bias how much and in what
    direction a participant changes their opinions. Failure to do so can result 
    in artificially-inflated effect sizes~\cite{Yarkoni2021}.
\end{itemize}

@@

Group polarization is often induced in behavioral experiments by first answering
an opinion prompt, grouping
relatively like-minded individuals together to discuss the prompt, then 
again polling opinions; the change in mean opinion towards a more extreme
consensus than the initial mean is the amount of group polarization that has occurred
(Figure~\ref{fig:groupPolarizationSchematic}).

While all ten studies scrutinized here use the general experimental procedure
outlined above (Figure~\ref{fig:groupPolarizationSchematic}), each study and constituent 
experimental condition have idosyncracies that makes the results themselves
more opaque, and that makes comparison of different studies' findings more
difficult. (REVIEW A FEW SIMPLE EXAMPLES). A detailed examination of each study's
idiosyncracies is given in Table~\ref{tab:conditionsComparison}.

\begin{figure}
  \caption{Group polarization experiments trials in the ten studies
    analyzed here proceed in three steps: 
    (A) \textbf{Pre-deliberation.} Participant opinions are measured using an 
    ordinal scale. 
    (B) \textbf{Deliberation.} Groups discuss their opinions on the prompt. 
    (C) \textbf{Post-deliberation.} Participant opinions are measured again.
    Group polarization in one trial is said to occur if 
    the mean post-deliberation opinion is greater than the pre-deliberation 
    opinion.
}
  \label{fig:groupPolarizationSchematic}
  \centering
  \includegraphics[width=\textwidth]{Figures/GPExperimentSketch.pdf}
\end{figure}

% \spacing{1.0}
% \input{tables/conditions_table.tex}
% \spacing{1.5}

@@

\section{Model}

Our model represnts the generative process of finding different forms of group consensus
in group polarization experiments, and the process of measuring that consensus using
an ordinal measurement scale, e.g.\ a Likert-type scale. This model operationalizes two
relevant forms of consensus: \emph{simple consensus} and \emph{group polarization}.
\mt{Need to explain in Intro why consensus should occur along with group
polarization} 
% Group polarization experiments are designed to result group consensus
% since they group together like-minded individuals to see if they become more
% extreme after discussion~\cite{Turner2018}.

We collected group polarization data from ten empirical studies with sixty constituent
experimental conditions across all studies. We recorded the reported opinion shift and pre- and
post-deliberation mean opinions, and the number of participants
in each condition, as long as this data was available---\citeA{Burnstein1975} 
did not report pre- and post-deliberation
means, only the shift in means. We did not include experimental treatments that did not 
yield a positive detection of group polarization since we are only estimating
the rate of plausible false positive detections of group polarization.

Based on this data

% In two cases, reported opinion and shift values were converted from an ordinal
% scale to a percentage scale, specifically \citeA{Friedkin1999a} and \citeA{Krizan2007}.
% The authors in these papers did report the original ordinal measurement scale,
% so we converted the means back to the orignial measurement scale.
% In a few cases from \citeA{Friedkin1999a} we did not need to apply our 
% model to the reported data because we could determine a reported opinion shift was 
% \emph{prima facie} false since reported opinion shift magnitudes
% were smaller than the standard deviation in shifts (see Table 1, p. 868, \emph{ibid}),
% despite Friedkin's claim that regressions (not clearly defined) on all shifts 
% yielded $p < 0.05$~\cite{Kruschke2018c}.

\subsection{Latent opinions, ordinal measurements, and how simple consensus can look like
group polarization}

We begin by developing a formal model of ordinal measurement of latent continuous
opinions (model variables and parameters are given in Table~\ref{tab:modelVariables}).
First, we assume that participant $i$ has latent opinion at time $t$ that is drawn from
a group-level normal distribution,
\begin{equation}
  o_{i,t} \sim p(o; \mu_t, \sigma_t) = \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution}
\end{equation}
\noindent 
This distribution has mean $\mu_t$ and standard deviation $\sigma_t$.
This form is chosen to
reflect the common statistical analysis used by group polarization studies,
not because it is statistically appropriate---in fact it is not, which is what
we aim to show in this paper.

It is appropriate to assume opinions are normally distributed because published
group polarization studies use some sort of $t$-tests, or similar tests, which are designed
to detect whether the mean of two distributions are the same or not. This imposes the
implicit assumption that opinions themselves are normal. But as we will show, 
metric statistical tests such as $t$-tests reliably generate false rejections
of the null hypothesis that there is no difference between two distributions' means.

\subsubsection{How simple consensus can look like group polarization}

Simple consensus can look like group polarization when (1) the mean latent opinion
does not change; (2) both more and less extreme opinions move toward this 
unchanged latent mean (i.e., opinion variance shrinks due to consensus); 
and (3) an ordinal scale is used that bins all extreme
latent opinions into the same extreme bin even though those means have changed,
but opinion bins that are less extreme or of opposite polarity more accurately
reflect the underlying change in latent opinion distribution. This process is
illustrated in Figure~\ref{fig:consensusDistroIllustration}, where we show
a toy example of what can occur when the pre-consensus
opinions (upper left plot)

\begin{table}[h]
  \caption{Variables used in the formal model of group polarization from simple consensus.}
  \label{tab:modelVariables}
  \begin{tabular}{cp{5.0in}} \toprule
   Symbol & Description  \\ \midrule  
   $t \in \{\mathrm{pre}, \mathrm{post}\}$ & Time, either \emph{pre}- or
                                           \emph{post}-discussion \\
   $\mu$  &   Latent mean, assumed constant for simple consensus \\ 
   $\sigma_t$ & Latent standard deviation at time $t$ \\
   $p(o;\mu,\sigma_t)$ & Normal latent distribution of opinions with mean
                         $\mu$ and variance $\sigma_t^2$ \\
   $k \in \{0,\ldots,B\} $ & Threshold index ($k$) between 0 and total number of
         ordinal bins, $B$ \\
   $b_k$ & Opinion bin $k$; note there is no $k=0$ bin. \\
   $\theta_{k-1},~\theta_k$  & Lower, upper threshold value for binning latent opinions into
     bin $b_k$ \\
   $p(o \in b_k; \mu, \sigma_t)$ & Ordinal density of binned opinions in bin $b_k$ 
            (Equation~\ref{eq:binFrequency}) \\
   $\langle o_t \rangle$  &   Observed mean opinion calculated from ordinal data at time $t$ \\
   \bottomrule
  \end{tabular} 
\end{table}

\begin{figure}
  \caption{\emph{Simple consensus} occurs in a group when its latent opinion
  distribution variance decreases, but the mean does not change (A). When 
latent opinions are measured on an ordinal scale, the mean of the distribution
can be distorted by ceiling (or floor) effects (B). Group polarization 
can appear to have occurred where the mean of the observed opinions appears
to increase despite the latent mean being constant (C).}
  \label{fig:consensusDistroIllustration}
  \vspace{1em}
  \centering
    \includegraphics[width=1.0\textwidth]{ConsensusDistroIllustration/figure.pdf}
\end{figure}

This distributional view models the measurement of $N \to \infty$ participant opinions,
which results in a histogram of frequency of opinions
in each bin, i.e., $o_{i,t} = b_k$. The frequency of responses in each bin is 
the integral over the continuous normal probability density function from 
one bin threshold to another. This transforms the probability density 
function from $p(o_{i,t};~\mu_t, \sigma_t)$ to the probability
of observing a reported opinion of each bin value. 
The probability of observing an opinion in bin
$b_k$ is calculated by integrating the normal probability density 
function over the range of $\theta_{k-1}$ to $\theta_{k}$. 
Formally, we write the 
probability of observing an opinion in bin $k$ at time $t$ as
\begin{equation}
\begin{aligned}
  p(o \in b_k; \mu_t, \sigma_t, \theta) 
    &= \int_{\theta_{k-1}}^{\theta_k} p(o;\mu_t, \sigma_t) do \\
    &= \Phi \left( \frac{o - \mu_t}{\sigma_t} \right)\Big|_{o=\theta_{k-1}}^{\theta_k} \\
    &= \Phi \left( \frac{\theta_k - \mu_t}{\sigma_t} \right) - 
       \Phi \left( \frac{\theta_{k-1} - \mu_t}{\sigma_t} \right)
\end{aligned}
\label{eq:binFrequency}
\end{equation}
\noindent
where $\Phi(\frac{o - \mu}{\sigma})$ is the normalized 
normal cumulative distribution function over opinions $o$, shifted by an amount $\mu$ with
standard deviation $\sigma$.

\subsection{Evaluating whether a group polarization finding is plausibly false}

We have shown in one toy example that group polarization can appear to occur
when ordinal measures are used, without accounting for the binning process; now
we show how to .

We develop two measures for how plausible it is that a reported positive detection
of group polarization is indeed a false positive, building our confidence 
in the plausibility at each step. In this way, we take a modern Bayesian approach
to scientific evidence that typically does not dichotomize ``true'' and ``false''
into absolutes, but evaluates degrees of belief instead.

The first measure we use for whether a reported detection of group polarization
is plausibly false is whether we can find a combination of latent simple
consensus parameters $\mu$, $\sigmapre$, and $\sigmapost$ representing latent
distributions that, when binned (Equation~\ref{eq:binFrequency} and
Figure~\ref{fig:consensusDistroIllustration}), result in ordinal distributions
that have the same difference in means reported in the literature that the authors
took to mean group polarization had occurred. Note that these calculations are
done for the case where effectively we have $N \to \infty$; next we turn to
the question of whether data generated from these simple consensus distributions
generate a ``significant'' group polarization result.

The second measure of how plausible it is that a group polarization is false
is whether ordinal data generated by binning draws from the latent distributions
is found to be significantly different according to Welch's $t$-test.

\subsubsection{Identify null-model parameters that yield plausibly-false positives}

To find simple consensus parameters $\mu$, $\sigmapre$, and $\sigmapost$ that
generate ordinal distributions showing group polarization, we used the above
framework in an optimization routine that takes an initial guess for these latent
parameters and hillclimbs to find $\sigmapre$ and $\sigmapost$ that most accurately
generate ordinal data that appears to have had group polarization occur.

\subsubsection{Metric model inference of Cohen's $d$ for simulated simple-consensus data}

After identifying null-model parameters that yield plausibly-false positives 
in the $N \to \infty$ case, we generated binned samples from normal opinion 
distributions $\normalpre$ and $\normalpost$. Specifically, for each experimental
condition that passed the initial $N \to \infty$ test, we generated 1000 
pre- and post-deliberation ordinal datasets.

We generated these simulated data using the latent mean, pre-deliberation 
standard deviation, and post-deliberation standard deviations as parameters of
normal distributions of pre- and post-deliberation opinions.

\subsection{Bayesian ordered-probit models inference of Cohen's $d$ for simulated
simple-consensus data}

Non-metric ordered probit models, which account for the data collection 
process, theoretically should be better able to correctly
distinguish simple consensus compared to metric models. Therefore, we also fit
the following ordered probit model to simulated group polarization data.

@@

\vspace{.5em}
\spacing{1.0}
% \input{plausibleFP-table.tex}
\begin{table}[H]
    \begin{center}
      \caption{Number and fraction of experimental conditions where group polarization was
               reported to have occurred, but plausibly did not.}
      \vspace{.5em}
      \input{plausibleFP-table-custom.tex}
    \end{center}
  \end{table}
\spacing{1.5}

\subsection{Metric linear models often fail to identify simple consensus}

Here we evaluate if metric linear models inferred group polarization from simulated
simple consensus data using parameters identified above. Those parameters specified
continous pre- and post-deliberation means and variance of normal opinion
distributions that represent simple consensus but yields
pre- and post-deliberation ordinal distributions with different, group-polarized
means. This parameter-finding process implicitly assumes $N\to\infty$ participants.

It is therefore necessary to evaluate whether the simple consensus parameters so 
identified in the previous step yield simulated data that also appear to be
group-polarized when measured on an ordinal scale. We draw $N$ samples from the
pre- and post-deliberation normal distributions and bin them into one of 
$B$ ordinal values. $N$ and $B$ are set specifically for each of the 
60 experimental conditions analyzed here. We then fit a linear model to the 
simulated observed pre- and post-deliberation data to infer whether the mean
shifted, indicating group polarization. We measure the pre- and
post-deliberation standard deviations as well. Then we calculate Cohen's $d$
to quantify the effect size for each simulation. A suite of simulations are
necessary since there might be significant variance in $d$ for different
generations of simulated pre- and post-deliberation observations. Therefore we
perform 1000 trials for each experimental condition and inspect the resulting
distributions of $d$ across trials in a boxplot.

\vspace{0.5em}
\begin{center}
{[ FIGURE \ref{fig:MetricBoxplot_N} ABOUT HERE ]} \\
\end{center}

\subsection{Ordered-probit models also fail?}

@@

\subsection{More data enables ordered probit models, but not metric models, to reliably
  identify simple consensus}

  \mt{Idea: increase $N$ for simulations and see effect on distribution of Cohen's $d$
  for both metric and ordinal models}

@@

We did make some simplifying assumptions that could affect model outcomes.

This problem is made all the worse by the lack of a consistent experimental design for group
polarization experiments. This makes it impossible to systematically account for
additional sources of vairance known to lead to over-estimates
of effect sizes~\cite{Yarkoni2021}, independent of the false positives problem
we identified in the current paper.

% This is due to the rush to explain why 
% group polarization occurs using weak theory~\cite{Cartwright1971,Cartwright1973} and makes
% group polarization meta-analyses (e.g. Brown, 1986; and Isenberg, 1986)\nocite{Brown1986,Isenberg1986}
% unreliable~\cite{Meehl1990,Meehl1997}.

### 2024-04-09 GroupPolarizationStatmod/paper/ms.qmd
init quarto paper rewrite

---
title: "Prominent claims of group polarization are plausibly false with lessons for studying psychological dynamics"
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    cite-method: biblatex
    biblatexoptions: sorting=none, style=numeric-comp
    pdf-engine: latexmk
    pdf-engine-opt: -xelatex
    # template-partials:
    #   - _extensions/partials/title.tex
authors:
  - name: Matthew Adam Turner
    institute: Stanford University Doerr School of Sustainability, Stanford University
    roles: writing, model development, project development
    corresponding: true
    email: maturner@stanford.edu
  - name: Paul E. Smaldino
    institute: 
      - Cognitive and Information Sciences, Univeristy of California, Merced
      - Santa Fe Institute
    roles: writing
    corresponding: true
bibliography: ../library.bib
classoption: abstract
keep-tex: true
abstract: |
  This is the abstract
geometry:
  - margin=1.1in
  - letterpaper
biblio-title: References
---

>>>>

---
title: "Prominent claims of group polarization are plausibly false with lessons for studying psychological dynamics"
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    cite-method: biblatex
    biblatexoptions: sorting=none, style=numeric-comp
    pdf-engine: latexmk
    pdf-engine-opt: -xelatex
    keep-tex: true
    template-partials:
      - _extensions/partials/title.tex
authors:
  - name: Matthew Adam Turner
    affiliation: Stanford University Doerr School of Sustainability, Stanford University
    roles: writing, model development, project development
    corresponding: true
    email: maturner@stanford.edu
  - name: Paul E. Smaldino
    affiliation: 
      - Cognitive and Information Sciences, Univeristy of California, Merced
      - Santa Fe Institute
    roles: writing
    corresponding: true
bibliography: ../library.bib
classoption: abstract
keep-tex: true
abstract: |
  This is the abstract
geometry:
  - margin=1.1in
  - letterpaper
biblio-title: References
---
